---
title: "Tutorial 4: Computer Exercises (Chapter 2) - C6, C7, C8"
date: "ECO375"
author: "Wenhui Xu"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
urlcolor: blue
geometry: left=0.75in, right=0.75in, top=0.75in, bottom=0.75in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```



# Overview

This tutorial covers three computer exercises from Wooldridge (2019), Chapter 2:

- **C6 (MEAP93):** a level-log model of math pass rates on spending per student.
- **C7 (CHARITY):** a simple regression of charitable giving on mailings per year.
- **C8 (Simulation):** generating data to illustrate sampling variation and OLS residual properties.

We will use base R as much as possible.

```{r packages}
# Install/load packages only if needed
if (!requireNamespace("wooldridge", quietly = TRUE)) install.packages("wooldridge")
library(wooldridge)

# Optional helpers (not strictly required)
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
library(dplyr)
```


# 0.5 A quick demonstration: the pipe operator `%>%`

In this course we will often use the **pipe operator** `%>%` (from `dplyr` / `magrittr`).  
Read it as **"then"**: it takes the object on the left and passes it into the function on the right.

## The basic rule

```{r pipe_basic}
# These two lines do the same thing:
sqrt(16)
16 %>% sqrt()
```

In general:

- `x %>% f()` means `f(x)`
- `x %>% f(a = 1)` means `f(x, a = 1)`

## A two-step pipeline

```{r pipe_demo_data}
demo_df <- data.frame(
  math10 = c(20, 40, 60),
  expend = c(3500, 4200, 5100)
)
```

```{r pipe_two_steps}
# Without pipes (read inside-out):
summary(dplyr::select(demo_df, math10, expend))

# With pipes (read top-down):
demo_df %>%
  dplyr::select(math10, expend) %>%
  summary()
```

## The placeholder `.` (when x is not the first argument)

Sometimes the function you want to call does *not* take your object as the first input.  
Then you can use `.` to mark where the left-hand object should go.

```{r pipe_dot_placeholder1}
v <- c(1, 2, 3, 4, 5)

# Same result:
paste(v, collapse = "-")
v %>% paste(collapse = "-")
```

```{r pipe_dot_placeholder2}
s <- "1-2-3-4-5"

# gsub(pattern, replacement, x)
gsub("-", "_", s)

# Using the dot placeholder:
s %>% gsub("-", "_", .)

```

## A common pattern in data analysis

"Take a dataset -> select columns -> create new variables -> summarize"

```{r pipe_common_pattern}
demo_df %>%
  dplyr::select(math10, expend) %>%
  dplyr::mutate(ln_expend = log(expend)) %>%
  summary()
```

> Note: `%>%` works because we load `dplyr` in this tutorial. If you ever see an error
> like "could not find function %>%", it means the package was not loaded.


# C6: MEAP93 (math10 vs log(expend))

## What the problem is asking (plain English)

1. Conceptually: does spending have diminishing returns for pass rates?
2. Show why, in the model math10 = b0 + b1 log(expend) + u, the quantity b1/10 is the change in **percentage points** in math10 when expend rises by 10%.
3. Estimate the regression and report the fitted equation with n and R^2.
4. Use the estimate to compute the impact of a 10% spending increase.
5. Check whether fitted values exceed 100, and explain why this is not a major concern here.

## Data and a quick look

```{r c6_load}
data("meap93")
meap93 <- as_tibble(meap93)

# Key variables
meap93 %>% select(math10, expend) %>% summary()
```

## (i) Diminishing effect intuition

A linear-in-levels model assumes each extra dollar has the same marginal effect.
A log specification allows the effect of spending to shrink as spending gets larger.

Here we proceed with the **level-log** model:
\[
math10 = b_0 + b_1 \log(expend) + u.
\]

## (ii) Why b1/10 corresponds to a 10% increase in expend

A 10% increase means expend_new = 1.10 * expend.
Then:

log(expend_new) - log(expend) = log(1.10) \approx 0.0953.

So the change in predicted math10 is approximately:

b1 * log(1.10) \approx b1 * 0.10 = b1/10.

Because math10 is measured in **percentage points**, this change is in percentage points too.

## (iii) Estimate the model and report the equation

```{r c6_reg}
c6_fit <- lm(math10 ~ log(expend), data = meap93)
summary(c6_fit)

b0_hat <- coef(c6_fit)[1]
b1_hat <- coef(c6_fit)[2]
n_c6   <- nobs(c6_fit)
r2_c6  <- summary(c6_fit)$r.squared

list(b0_hat = b0_hat, b1_hat = b1_hat, n = n_c6, r2 = r2_c6)
```

Write the fitted equation:

```{r c6_equation, results='asis'}
cat(sprintf("Estimated equation:\n\nmath10 = %.3f + %.3f * log(expend)\n\nn = %d,   R^2 = %.4f\n",
            b0_hat, b1_hat, n_c6, r2_c6))
```

## (iv) Effect of a 10% increase in spending

```{r c6_effect10}
effect_10pct_exact <- b1_hat * log(1.10)
effect_10pct_approx <- b1_hat / 10

c(effect_10pct_exact = effect_10pct_exact,
  effect_10pct_approx = effect_10pct_approx)
```

Interpretation: a 10% increase in spending is associated with an increase of about the number above (in **percentage points**) in the math pass rate.

## (v) Can fitted values exceed 100?

```{r c6_fitted}
meap93 <- meap93 %>% mutate(math10_hat = fitted(c6_fit))

summary(meap93$math10_hat)
sum(meap93$math10_hat > 100)
max(meap93$math10_hat)
```

Why this is not much of a worry here:

- The sample range of log(expend) is limited, so predicted values tend to stay in a realistic range.
- Even if a few fitted values were slightly above 100, that is a limitation of the linear regression functional form, not necessarily a data problem.

# C7: CHARITY (gift vs mailsyear)

## What the problem is asking (plain English)

1. Compute average gift and the percentage of people who gave zero.
2. Summarize mailings per year: mean, min, max.
3. Estimate gift = b0 + b1 mailsyear + u and report n and R^2.
4. Interpret b1 and discuss net gain if each mailing costs one guilder.
5. Find the smallest predicted gift. Can the model ever predict exactly zero?

## Data checks

```{r c7_load}
data("charity")
charity <- as_tibble(charity)

charity %>% select(gift, mailsyear) %>% summary()
```

## (i) Average gift and percent zero

```{r c7_part_i}
avg_gift <- mean(charity$gift, na.rm = TRUE)
pct_zero <- mean(charity$gift == 0, na.rm = TRUE) * 100

c(avg_gift = avg_gift, pct_zero = pct_zero)
```

## (ii) Mailings per year: mean, min, max

```{r c7_part_ii}
c(mean_mailsyear = mean(charity$mailsyear, na.rm = TRUE),
  min_mailsyear  = min(charity$mailsyear, na.rm = TRUE),
  max_mailsyear  = max(charity$mailsyear, na.rm = TRUE))
```

## (iii) Estimate the regression and report

```{r c7_reg}
c7_fit <- lm(gift ~ mailsyear, data = charity)
summary(c7_fit)

b0_c7 <- coef(c7_fit)[1]
b1_c7 <- coef(c7_fit)[2]
n_c7  <- nobs(c7_fit)
r2_c7 <- summary(c7_fit)$r.squared

list(b0_hat = b0_c7, b1_hat = b1_c7, n = n_c7, r2 = r2_c7)
```

```{r c7_equation, results='asis'}
cat(sprintf("Estimated equation:\n\ngift = %.3f + %.3f * mailsyear\n\nn = %d,   R^2 = %.4f\n",
            b0_c7, b1_c7, n_c7, r2_c7))
```

## (iv) Interpret b1 and net gain per mailing

- b1 is the **expected change in gift (guilders)** when mailings per year increases by 1.
- If a mailing costs 1 guilder, then the expected net gain from one additional mailing is approximately (b1 - 1).

```{r c7_netgain}
expected_net_gain_per_mailing <- b1_c7 - 1
expected_net_gain_per_mailing
```

Important nuance: this is an **average** effect across the sample, not a guarantee that every mailing to every person produces a gain.

## (v) Smallest predicted gift and can we predict zero?

```{r c7_pred}
charity <- charity %>% mutate(gift_hat = fitted(c7_fit))

min_pred <- min(charity$gift_hat, na.rm = TRUE)
min_pred
```

Can the simple regression ever predict zero?

- The fitted value is: gift_hat = b0_hat + b1_hat * mailsyear.
- This equals zero only if mailsyear = -b0_hat / b1_hat, which may not be an integer and may be outside the observed range.
- If b0_hat > 0 and mailsyear is nonnegative, then predictions will tend to be positive.

# C8: Simulation and OLS properties

## What the problem is asking (plain English)

We simulate data from a known population process:

1. Draw x from Uniform(0, 10) and compute sample mean and sd.
2. Draw u from Normal(0, 36) and compute sample mean and sd; explain why sample mean is not exactly zero.
3. Generate y = 1 + 2x + u, run OLS, and compare estimates to the true values (1 and 2).
4. Compute residuals and verify the OLS residual properties: sum(uhat) = 0 and sum(x*uhat) = 0 (up to rounding).
5. Replace residuals by the *true* errors u and see these properties do not hold in finite samples.
6. Repeat (i)-(iii) with a new sample and explain why estimates change.

## A clean, reproducible implementation

```{r c8_simulation}
set.seed(375)  # reproducibility

n <- 500

# (i) x ~ Uniform(0,10)
x <- runif(n, min = 0, max = 10)
mean_x <- mean(x)
sd_x   <- sd(x)

# (ii) u ~ Normal(0,36) so sd = 6
u <- rnorm(n, mean = 0, sd = 6)
mean_u <- mean(u)
sd_u   <- sd(u)

# (iii) y = 1 + 2x + u
y <- 1 + 2*x + u

fit <- lm(y ~ x)

b0_hat <- coef(fit)[1]
b1_hat <- coef(fit)[2]

list(mean_x = mean_x, sd_x = sd_x,
     mean_u = mean_u, sd_u = sd_u,
     b0_hat = b0_hat, b1_hat = b1_hat)
```

### (ii) Why is the sample mean of u not exactly zero?

Even though E[u] = 0 in the population, a finite random sample will almost never have sample mean exactly 0.
As n grows, the sample mean tends to get closer to 0 (law of large numbers).

### (iv) OLS residual properties (sample analog of equation 2.60)

```{r c8_residual_properties}
uhat <- resid(fit)

# Theoretical properties (in sample, with intercept):
# sum(uhat) = 0 and sum(x * uhat) = 0, up to rounding error.
c(sum_uhat = sum(uhat),
  sum_x_uhat = sum(x * uhat))
```

### (v) Same computation using the *true* errors u

```{r c8_true_errors}
c(sum_u = sum(u),
  sum_x_u = sum(x * u))
```

Conclusion: the residual properties are guaranteed by the OLS algebra, but the same equalities do not generally hold if we plug in the unobserved true errors u.

### (vi) Repeat (i)-(iii) with a new sample

```{r c8_repeat}
set.seed(376)  # new seed -> new sample

x2 <- runif(n, 0, 10)
u2 <- rnorm(n, 0, 6)
y2 <- 1 + 2*x2 + u2

fit2 <- lm(y2 ~ x2)
coef(fit2)
```

Why are the new estimates different?

Because each sample has different random draws of x and u, so the OLS estimates vary from sample to sample. With larger n, they concentrate around the true values (consistency).

# Wrap-up

- C6 shows how to interpret a level-log model and translate log changes into percentage changes.
- C7 shows how to interpret a simple regression, including the difference between an average effect and an individual outcome.
- C8 connects probability theory to the mechanics of OLS: residual properties hold by construction, and sampling variation is unavoidable.
