---
title: "Tutorial 6: Multiple Regression + Inference (Chapter 3 C2 & Chapter 4 C1)"
date: "ECO375"
author: "Wenhui Xu"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
urlcolor: blue
geometry: left=0.75in, right=0.75in, top=0.75in, bottom=0.75in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This tutorial covers:

- **Chapter 3, Computer Exercise C2 (HPRICE1):** multiple regression, prediction, and residuals.
- **Chapter 4, Computer Exercise C1 (VOTE1):** log-level model and testing a linear restriction.
- Two supporting tools (used in the original script):
  - VIF (variance inflation factor),
  - Linear hypothesis testing (F tests).

We will use the pipe operator `%>%` (from `dplyr`) for readability.

```{r packages}
if (!requireNamespace("wooldridge", quietly = TRUE)) install.packages("wooldridge")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("car", quietly = TRUE)) install.packages("car")

library(wooldridge)
library(dplyr)
library(car)
```

# 1. Supporting tools (quick)

## 1.1 VIF (Variance Inflation Factor)

VIF for regressor j is:
\[
VIF_j = \frac{1}{1 - R_j^2},
\]
where \(R_j^2\) is from regressing \(x_j\) on the other regressors.

```{r vif_demo}
data("wage1")

lm_wage1 <- lm(log(wage) ~ educ + exper + tenure, data = wage1)

# VIFs
vif(lm_wage1)

# Example: compute R^2_j for educ and confirm VIF formula
r2_educ <- summary(lm(educ ~ exper + tenure, data = wage1))$r.squared
c(R2_educ = r2_educ, VIF_educ = 1/(1 - r2_educ))
```

## 1.2 Linear hypothesis tests (F-tests)

A joint hypothesis like
\[
H_0: \beta_1 = 0, \; \beta_2 = 0, \; \beta_3 = 0
\]
is tested using an F-test.

```{r ftest_demo}
linearHypothesis(lm_wage1, c("educ = 0", "exper = 0", "tenure = 0"))
```

# 2. Chapter 3 — Computer Exercise C2 (HPRICE1)

## 2.1 What the problem is asking 

We estimate:
\[
price = \beta_0 + \beta_1\,sqrft + \beta_2\,bdrms + u,
\]
where **price is measured in thousands of dollars**.

Tasks:
1. Estimate the regression and write it in equation form.
2. Interpret \(\hat\beta_2\): one more bedroom holding sqrft constant.
3. Compute the predicted price change for adding a bedroom **and** 140 square feet.
4. Report the percent variation in price explained by sqrft and bdrms (use \(R^2\)).
5. Predict the selling price for the first house in the sample.
6. Compute the residual for the first house and interpret underpaid vs overpaid.

## 2.2 Step-by-step plan (how to solve it)

1) Load `hprice1` and run `lm(price ~ sqrft + bdrms)`.  
2) Extract coefficients, standard errors, n, and \(R^2\).  
3) Keep track of units: changes in `price` are in **thousands of dollars** (multiply by 1000 for dollars).  
4) Use linear prediction:
\[
\Delta \hat{price} = 140\,\hat\beta_1 + 1\cdot\hat\beta_2.
\]
5) Use `predict()` for the first house. Residual = actual − predicted.

## 2.3 Common pitfalls

- **Units mistake:** `price` is in thousands of dollars.
- **R-squared choice:** the question asks “percent explained,” which is usually \(R^2\), not adjusted \(R^2\).
- **Interpretation slip:** \(\hat\beta_2\) is “one more bedroom holding sqrft fixed.”

## 2.4 Solution in R

```{r c2_load}
data("hprice1")
hprice1 <- as_tibble(hprice1)

hprice1 %>% select(price, sqrft, bdrms) %>% summary()
```

```{r c2_regression}
fit_c2 <- lm(price ~ sqrft + bdrms, data = hprice1)
summary(fit_c2)

coef_c2 <- coef(fit_c2)
se_c2 <- sqrt(diag(vcov(fit_c2)))
n_c2 <- nobs(fit_c2)
r2_c2 <- summary(fit_c2)$r.squared

list(coef = coef_c2, se = se_c2, n = n_c2, r2 = r2_c2)
```

```{r c2_equation, results='asis'}
cat(sprintf(
  "Estimated equation (SE in parentheses):\n\nprice = %.3f (%.3f) + %.5f (%.5f)*sqrft + %.3f (%.3f)*bdrms\n\nn = %d,   R^2 = %.4f\n",
  coef_c2[1], se_c2[1],
  coef_c2[2], se_c2[2],
  coef_c2[3], se_c2[3],
  n_c2, r2_c2
))
```

### (ii) One more bedroom (holding sqrft constant)

```{r c2_part_ii}
delta_one_bed_thousands <- coef_c2["bdrms"]
delta_one_bed_dollars <- 1000 * delta_one_bed_thousands
delta_one_bed_dollars
```

### (iii) One more bedroom + 140 square feet

```{r c2_part_iii}
delta_thousands <- 140 * coef_c2["sqrft"] + coef_c2["bdrms"]
delta_dollars <- 1000 * delta_thousands
delta_dollars
```

### (iv) Percent variation explained (R-squared)

```{r c2_part_iv}
r2_c2
```

### (v) Predicted selling price for the first house

```{r c2_part_v}
first_house <- hprice1[1, ]
pred_first_thousands <- predict(fit_c2, newdata = first_house)
pred_first_dollars <- 1000 * pred_first_thousands
pred_first_dollars
```

### (vi) Residual for the first house; underpaid or overpaid?

```{r c2_part_vi}
actual_first_thousands <- first_house$price
resid_first_thousands <- actual_first_thousands - pred_first_thousands
resid_first_dollars <- 1000 * resid_first_thousands

c(resid_thousands = resid_first_thousands, resid_dollars = resid_first_dollars)
```

Interpretation:
- If residual < 0: actual price < predicted price → buyer paid less than predicted (underpaid relative to observables).
- If residual > 0: buyer paid more than predicted (overpaid relative to observables).

# 3. Chapter 4 — Computer Exercise C1 (VOTE1)

## 3.1 What the problem is asking (plain English)

We estimate:
\[
voteA = \beta_0 + \beta_1 \log(expendA) + \beta_2 \log(expendB) + \beta_3 prtystrA + u.
\]

Tasks:
1. Interpret \(\beta_1\).
2. State the null hypothesis that a 1% increase in A's spending is offset by a 1% increase in B's spending.
3. Estimate the model and discuss whether A and B spending affect the outcome.
4. Estimate a model that directly gives the t statistic for testing the null in (2) (two-sided).

## 3.2 Step-by-step plan (how to solve it)

1) Fit the model with OLS:
`lm(voteA ~ log(expendA) + log(expendB) + prtystrA)`.

2) Interpret log-level coefficients:
- A 1% increase in spending changes log(spending) by about 0.01.
- So \(\beta_1/100\) is approximately the percentage-point change in `voteA` from a 1% increase in `expendA`.

3) Offset null:
- “Offset” means the two 1% changes cancel:
\[
\beta_1 + \beta_2 = 0.
\]

4) Test properly:
- Use `linearHypothesis()` to test \(\beta_1+\beta_2=0\), **or**
- Reparameterize so one coefficient equals \(\beta_1+\beta_2\) and read its t statistic.

## 3.3 Common pitfalls

- **Log interpretation mistakes:** coefficients are tied to percentage changes, not dollar changes.
- **Testing incorrectly:** you need the standard error of \(\hat\beta_1+\hat\beta_2\).
- **Wrong-sided test:** part (iv) is two-sided.

## 3.4 Solution in R

```{r c1_load}
data("vote1")
vote1 <- as_tibble(vote1)

vote1 %>% select(voteA, expendA, expendB, prtystrA) %>% summary()
```

```{r c1_model}
fit_vote <- lm(voteA ~ log(expendA) + log(expendB) + prtystrA, data = vote1)
summary(fit_vote)
```

### (ii) State and test the offset null: \(\beta_1 + \beta_2 = 0\)

```{r c1_test_linearhypothesis}
linearHypothesis(fit_vote, "log(expendA) + log(expendB) = 0")
```

### (iv) Reparameterize to get the t statistic directly

Define:
- \(z_1 = \log(expendA) - \log(expendB)\)
- \(z_2 = \log(expendB)\)

Then:
\[
\beta_1 \log(expendA) + \beta_2 \log(expendB)
= \beta_1 z_1 + (\beta_1 + \beta_2) z_2.
\]

So the coefficient on \(z_2\) equals \(\beta_1 + \beta_2\).

```{r c1_reparam}
fit_vote_reparam <- lm(
  voteA ~ I(log(expendA) - log(expendB)) + log(expendB) + prtystrA,
  data = vote1
)

summary(fit_vote_reparam)
```

# Wrap-up

- **Ch.3 C2:** multiple regression, units, prediction, residual interpretation.
- **Ch.4 C1:** testing linear restrictions correctly (offset hypothesis).
- **Tools:** VIF (diagnostic), `linearHypothesis()` (clean restriction tests).
