---
title: "Tutorial 10: Instrumental Variables and 2SLS (MROZ) — v2"
date: "ECO375"
author: "Wenhui Xu"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
urlcolor: blue
geometry: left=0.75in, right=0.75in, top=0.75in, bottom=0.75in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This tutorial builds an end-to-end IV workflow (2SLS) using the **MROZ** dataset:

1) How to estimate **2SLS in R** (`AER::ivreg`)  
2) How to report **correct standard errors** (including **robust** IV standard errors)  
3) How to run a **Hausman-type endogeneity test** (Durbin–Wu–Hausman / Wu–Hausman)  
4) How to assess **instrument strength** (first-stage F statistics; “weak IV” diagnostics)

We use pipes (`%>%`) for clarity.

```{r packages}
if (!requireNamespace("wooldridge", quietly = TRUE)) install.packages("wooldridge")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("AER", quietly = TRUE)) install.packages("AER")

# For robust standard errors and formatted tests:
if (!requireNamespace("sandwich", quietly = TRUE)) install.packages("sandwich")
if (!requireNamespace("lmtest", quietly = TRUE)) install.packages("lmtest")

# For F-tests on excluded instruments:
if (!requireNamespace("car", quietly = TRUE)) install.packages("car")

library(wooldridge)
library(dplyr)
library(AER)
library(sandwich)
library(lmtest)
library(car)
```

# Module 1: What is `AER::ivreg()`? (first-time tool)

`ivreg()` fits an **instrumental variables** regression (2SLS under the hood).

## Why do we need IV / 2SLS?

OLS requires the regressor to be exogenous:

\[
E(u \mid x) = 0.
\]

If `educ` is correlated with the error term in the wage equation (e.g., ability/omitted factors), OLS is biased.  
An **instrument** \(z\) must satisfy:

1. **Relevance:** \(\mathrm{Cov}(z, x) \neq 0\)  
2. **Exogeneity:** \(\mathrm{Cov}(z, u) = 0\)

## `ivreg()` formula syntax

```r
ivreg(y ~ x1 + x2 + ... | z1 + z2 + ...)
```

- Left of `|` = regressors in the structural equation (endogenous + exogenous)
- Right of `|` = instruments (must include all exogenous regressors as their own instruments, plus excluded instruments)

Example (one endogenous regressor `educ`, one instrument `fatheduc`):

```r
ivreg(lwage ~ educ | fatheduc)
```

# 1. Data: MROZ working women (non-missing wages)

## What the problem is asking

We want the **causal** effect of education on log wage. We will:

1. Estimate the OLS return to education.
2. Re-estimate using IV, treating `educ` as endogenous.
3. Compare OLS vs IV estimates.
4. Understand the IV estimator in the just-identified case (covariance ratio).
5. Understand 2SLS (first stage + second stage) and why naive second-stage SE are wrong.
6. Estimate a richer 2SLS model with controls and multiple instruments.
7. Add diagnostics: robust IV SE, Hausman-type endogeneity test, and first-stage strength checks.

## Step-by-step plan

1) Load `mroz` and keep observations with non-missing `lwage`.  
2) Run OLS: `lm(lwage ~ educ)`.  
3) Run IV: `ivreg(lwage ~ educ | fatheduc)`.  
4) Compute IV slope manually: \(\widehat\beta_1 = \mathrm{Cov}(z,y)/\mathrm{Cov}(z,x)\).  
5) Do 2SLS by hand: first stage `educ ~ fatheduc`, second stage `lwage ~ educhat` (for intuition).  
6) Run the richer model:
\[
lwage = \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 exper^2 + u,
\]
instrumenting `educ` with parents’ education.  
7) Add **robust IV SE**, **Hausman-type test**, and **first-stage F** strength checks.

## Common pitfalls

- **Instrument validity:** relevance is not enough; exogeneity is the hard part (cannot be fully tested).
- **Naive 2SLS standard errors:** `lm(lwage ~ educhat)` reports invalid SE.
- **Forgetting exogenous controls must appear on both sides of `|`** in `ivreg()`.
- **Log wages interpretation:** 0.10 in log points is roughly a 10% wage change (approx.).

## Load and filter

```{r data_load}
data("mroz")

df <- mroz %>%
  as_tibble() %>%
  filter(!is.na(lwage))

df %>% summarise(n = n())
df %>% select(lwage, educ, exper, fatheduc, motheduc) %>% summary()
```

# 2. Baseline OLS: return to education

```{r ols}
ols1 <- lm(lwage ~ educ, data = df)
summary(ols1)
```

# 3. IV (just-identified): instrument `educ` with `fatheduc`

```{r iv_just_identified}
iv1 <- ivreg(lwage ~ educ | fatheduc, data = df)
summary(iv1)
```

# 4. Manual IV estimator (covariance ratio)

In a just-identified model with one instrument \(z\) and one endogenous regressor \(x\), the IV slope is:

\[
\widehat\beta_{1,IV} = \frac{\widehat{\mathrm{Cov}}(z,y)}{\widehat{\mathrm{Cov}}(z,x)}.
\]

```{r manual_iv}
beta1hat_iv <- with(df, cov(fatheduc, lwage) / cov(fatheduc, educ))
beta0hat_iv <- with(df, mean(lwage) - beta1hat_iv * mean(educ))

c(beta0hat_iv = beta0hat_iv, beta1hat_iv = beta1hat_iv)
coef(iv1)
```

# 5. Manual 2SLS (intuition) and why naive SE are wrong

## 5.1 First stage

```{r first_stage}
fs <- lm(educ ~ fatheduc, data = df)
summary(fs)

df <- df %>% mutate(educhat = fitted(fs))
```

## 5.2 Second stage (naive)

```{r second_stage_naive}
ss_naive <- lm(lwage ~ educhat, data = df)
summary(ss_naive)
```

**Key point:** the coefficient is typically close to IV, but the **standard errors are not valid** for 2SLS because `educhat` is estimated. Use `ivreg()` for correct 2SLS inference.

# 6. Richer 2SLS model (controls + multiple instruments)

We estimate:

\[
lwage = \beta_0 + \beta_1 educ + \beta_2 exper + \beta_3 exper^2 + u,
\]

treating `educ` as endogenous, and using `fatheduc` and `motheduc` as excluded instruments.

## Module 2: The `I()` function in formulas (first-time tool in this doc)

In R formulas, `I()` means “interpret this literally as arithmetic.”  
For example, `I(exper^2)` includes the variable \(exper^2\) in the regression.

```{r iv_richer}
iv2 <- ivreg(
  lwage ~ educ + exper + I(exper^2) |
    exper + I(exper^2) + fatheduc + motheduc,
  data = df
)

summary(iv2)
```

# 7. Compare OLS vs IV education slopes

```{r compare_slopes}
c(
  ols_educ = coef(ols1)["educ"],
  iv1_educ = coef(iv1)["educ"],
  iv2_educ = coef(iv2)["educ"]
)
```

# 8. Getting the "correct" standard errors in IV

## 8.1 What the problem is asking 

You need standard errors that are valid for IV/2SLS. Two practical layers:

1) **Correct 2SLS SE**: do not use the naive second-stage SE; use `ivreg()` output.  
2) **Robust 2SLS SE** (recommended in applied work): allow heteroskedasticity using `sandwich`.

## 8.2 Step-by-step plan

1) Use `summary(iv2)` for conventional (homoskedastic) 2SLS SE.  
2) Compute robust VCOV with `sandwich::vcovHC(iv2, type="HC1")`.  
3) Print coefficient tests with robust SE using `lmtest::coeftest()`.

## 8.3 Common pitfalls

- Robust SE change **t statistics/p-values**, not coefficients.
- HC0 vs HC1: HC1 is a common small-sample correction.
- Don’t mix OLS robust SE with IV models: use `vcovHC()` on the **ivreg object**.

## How to read the two outputs (conventional IV SE vs robust IV SE)

You will often see two slightly different standard errors for the *same* IV coefficient estimates:

- `summary(iv2)` reports **conventional 2SLS standard errors**, which rely on the homoskedasticity assumption.
- `coeftest(iv2, vcov. = vcovHC(iv2, ...))` reports **robust (heteroskedasticity-consistent) IV standard errors**.

**What is being compared?**  
Not “IV vs non-IV.” The coefficient estimates \(\hat\beta\) are the same IV/2SLS estimates in both cases.  
What changes is the **variance estimator** used for inference (SE, t-stat, p-value).

**Which is more accurate?**
- If the error variance is truly constant (homoskedastic), conventional SE can be more efficient (often smaller).
- If heteroskedasticity is present (common in practice), robust SE are typically more reliable for inference.

A common applied default is to **report robust IV SE (HC1)** unless the assignment explicitly asks for the homoskedastic version.


## Module 3: What are `vcovHC()` and `coeftest()`? (first-time tools)

### `sandwich::vcovHC()`
Computes a heteroskedasticity-consistent covariance matrix for coefficient estimates.  
You can apply it to `lm` and also to `ivreg` objects.

### `lmtest::coeftest()`
Prints coefficient tests using a user-supplied covariance matrix (e.g., robust VCOV).

## 8.4 Robust IV standard errors (HC1)

```{r robust_iv_se}
robust_vcov_iv2 <- vcovHC(iv2, type = "HC1")

# Robust coefficient table (large-sample reference by default)
coeftest(iv2, vcov. = robust_vcov_iv2)
```

Optional: compare conventional vs robust SE:

```{r compare_se}
se_conventional <- sqrt(diag(vcov(iv2)))
se_robust <- sqrt(diag(robust_vcov_iv2))

cbind(se_conventional = se_conventional, se_robust = se_robust)
```

# 9. Hausman-type test for endogeneity (DWH / Wu–Hausman)

## 9.1 What the problem is asking 

We want to test whether `educ` behaves as **endogenous** in the wage equation.

A Hausman-type test compares OLS and IV in a structured way. In `ivreg`, this is available as a **Wu–Hausman** (Durbin–Wu–Hausman) diagnostic.

## 9.2 Step-by-step plan

1) Estimate `iv2`.  
2) Call `summary(iv2, diagnostics = TRUE)`.  
3) Read the **Wu–Hausman** test:
- If p-value is small: evidence that OLS is inconsistent (endogeneity), IV preferred.
- If p-value is large: no strong evidence against exogeneity (OLS may be OK).

## 9.3 Common pitfalls

- This tests **endogeneity of the instrumented regressor(s)**, not instrument validity.
- A weak instrument can make IV noisy; endogeneity tests can have low power.
- “Fail to reject” does not prove exogeneity; it means insufficient evidence.

## Module 4: What does `summary(ivreg_model, diagnostics = TRUE)` do?

For `ivreg` objects, setting `diagnostics = TRUE` prints standard IV diagnostics, typically including:

- **Weak instruments** (first-stage relevance, reported via an F-type statistic)
- **Wu–Hausman** (endogeneity / Hausman-type test)
- **Sargan** (overidentification test when overidentified; relies on homoskedasticity)

```{r iv_diagnostics}
summary(iv2, diagnostics = TRUE)
```

# 10. Checking IV strength for **iv2 (MROZ)**: first-stage F statistics

In the richer MROZ IV model (`iv2`), we treat **educ** as endogenous and use **parents’ education** as excluded instruments:

- Endogenous regressor: `educ`
- Excluded instruments: `fatheduc`, `motheduc`
- Exogenous controls (also included as instruments): `exper`, `exper^2`

So the relevant first stage for **iv2** is:

\[
educ = \pi_0 + \pi_1 exper + \pi_2 exper^2 + \pi_3 fatheduc + \pi_4 motheduc + v.
\]

What we care about is whether the **excluded instruments** explain variation in `educ` *after controlling for* `exper` and `exper^2`.

## 10.1 What the problem is asking

We want evidence that the excluded instruments (`fatheduc`, `motheduc`) are correlated with `educ` **conditional on** the exogenous controls.

A standard check is the **joint F test** in the first stage:

\[
H_0:\ \pi_3 = 0,\ \pi_4 = 0.
\]

This is the “instrument strength” check most often referenced by the rule of thumb “first-stage F > 10” (very rough guideline).

## 10.2 Step-by-step plan

1) Fit the first-stage regression used by `iv2`:
\[
educ = \pi_0 + \pi_1 exper + \pi_2 exper^2 + \pi_3 fatheduc + \pi_4 motheduc + v.
\]
2) Test the joint null that the **excluded instruments** have zero coefficients:
\[
H_0: \pi_3 = 0,\ \pi_4 = 0.
\]
3) Report the resulting F statistic and p-value.
4) Interpret: larger F means stronger partial correlation between instruments and `educ`.

## 10.3 Common pitfalls

- Using the overall model F instead of the **excluded-instruments** joint F.
- Forgetting to include exogenous controls in the first stage when they appear in the second stage.
- Treating “F > 10” as a guarantee (it is not; it is a rule of thumb).

## Module 5: What is `car::linearHypothesis()`? (first-time tool in this doc)

`linearHypothesis()` conducts Wald/F tests of linear restrictions like:
\[
H_0:\ \pi_3 = 0,\ \pi_4 = 0.
\]

It is an easy way to obtain the **joint F test** for excluded instruments.

## 10.4 First stage + excluded instruments joint F

```{r first_stage_strength}
fs2 <- lm(educ ~ exper + I(exper^2) + fatheduc + motheduc, data = df)
summary(fs2)

# Joint F test: excluded instruments are irrelevant
linearHypothesis(fs2, c("fatheduc = 0", "motheduc = 0"))
```

# Wrap-up

- **2SLS in R:** `AER::ivreg()`
- **Correct SE:** use `ivreg()` (not naive second stage); add robust SE via `vcovHC()` + `coeftest()`
- **Hausman-type test:** `summary(iv2, diagnostics = TRUE)` provides Wu–Hausman (endogeneity)
- **Instrument strength:** check excluded-instruments joint F in the first stage (and the weak-instruments diagnostic)
