---
title: "Tutorial 3 - t-tests, Confidence Intervals, and the OLS Estimator (with Simulation)"
date: "ECO375"
author: "Wenhui Xu"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
urlcolor: blue
geometry: left=0.75in, right=0.75in, top=0.75in, bottom=0.75in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 0. Setup

This tutorial builds two "muscle groups" that show up everywhere in econometrics:

1) **One-sample t tests** (hypotheses, critical values, p-values, confidence intervals)  
2) **Simple OLS regression** (run with `lm()`, then replicate coefficients using formulas)

Datasets come from the `wooldridge` package:
- `audit` (Urban Institute audit study example; variable `y = b - w`)
- `ceosal1` (CEO salary vs. ROE example)

```{r}
# Install once if needed:
# install.packages("wooldridge")

library(wooldridge)

data("audit")
data("ceosal1")
```

# 1. One-sample t test in R

## 1.1 Warm-up: t.test() on simulated data
```{r}
set.seed(1)
x <- rnorm(100, mean = 0, sd = 1)
```

```{r}

# Two-sided test of H0: mu = 0 at alpha = 0.05 
# we know, sampling data X, are drawn from a distribution with true parameter mu_x = 0, sqrt(Var_x)=sd_x =1
t.test(x)

#t = 1.2123 <- test-statistic = (x_bar - mu_x )/se_x 
#for two sided, reject if |t-statistic| >= Critical_value_0.05
# 1.2123 < 1.96, fail to reject.
# df = n-1
# reject if p-value is smaller than alpha(5%)
# CI x_hat +- t-statistic * critical value

# One-sided test of H0: mu = 0 vs H1: mu < 0
t.test(x, alternative = "less")

# One-sided test with alpha = 0.10 (=> 90% CI)
t.test(x, alternative = "less", conf.level = 0.90)

# Test H0: mu = 4 (just to see the syntax)
t.test(x, mu = 4)
```

## 1.2 DIY t test (to understand what t.test() is doing)

### What we want our function to do
Inputs:  
- `x`: numeric sample  
- `mu0`: hypothesized mean under H0  
- `alpha`: significance level  
- `alternative`: `"two.sided"`, `"less"`, or `"greater"`  

Outputs (at least):  
- t statistic, df, critical value, p-value, confidence interval, and a decision.

```{r}
my_ttest <- function(x, mu0 = 0, alpha = 0.05,
                     alternative = c("two.sided", "less", "greater")) {

  alternative <- match.arg(alternative)
  x <- as.numeric(x)
  n <- length(x)
  if (n < 2) stop("Need at least 2 observations.")

  xbar <- mean(x)
  s <- sd(x)
  se <- s / sqrt(n)
  df <- n - 1

  t_stat <- (xbar - mu0) / se

  if (alternative == "two.sided") {
    crit <- qt(1 - alpha/2, df)
    p_value <- 2 * (1 - pt(abs(t_stat), df))
    CI <- c(xbar - crit * se, xbar + crit * se)
    decision <- if (abs(t_stat) > crit) "Reject H0" else "Fail to reject H0"

  } else if (alternative == "less") {
    crit <- qt(alpha, df)                 # left-tail critical value
    p_value <- pt(t_stat, df)             # left-tail probability
    CI <- c(-Inf, xbar + qt(1 - alpha, df) * se)  # one-sided CI (upper bound)
    decision <- if (t_stat < crit) "Reject H0" else "Fail to reject H0"

  } else { # alternative == "greater"
    crit <- qt(1 - alpha, df)             # right-tail critical value
    p_value <- 1 - pt(t_stat, df)         # right-tail probability
    CI <- c(xbar - qt(1 - alpha, df) * se, Inf)  # one-sided CI (lower bound)
    decision <- if (t_stat > crit) "Reject H0" else "Fail to reject H0"
  }

  list(
    n = n,
    df = df,
    xbar = xbar,
    s = s,
    se = se,
    t = t_stat,
    crit_value = crit,
    p_value = p_value,
    CI = CI,
    alternative = alternative,
    alpha = alpha,
    mu0 = mu0,
    decision = decision
  )
}

# quick check vs built-in
my_ttest(x, mu0 = 0, alpha = 0.05, alternative = "two.sided")
t.test(x, mu = 0, alternative = "two.sided")
```

# 2. Simulation: does the t statistic behave like a t distribution under H0?

## Goal
If the sample really comes from a normal distribution with mean mu0, then the one-sample t statistic
\[
t = \frac{\bar x - \mu_0}{s/\sqrt{n}}
\]
follows a \(t_{n-1}\) distribution.

We'll check this by simulation.

```{r}
set.seed(123)

R <- 5000
n <- 10
mu0 <- 1
sigma <- 2
alpha <- 0.07

t_stats <- numeric(R)
reject <- logical(R)

for (r in 1:R) {
  x_sim <- rnorm(n, mean = mu0, sd = sigma)  # H0 is true
  out <- my_ttest(x_sim, mu0 = mu0, alpha = alpha, alternative = "two.sided")
  t_stats[r] <- out$t
  reject[r] <- (out$decision == "Reject H0")
}

# Compare histogram to theoretical t density
hist(t_stats, probability = TRUE,
     main = "Simulated t-statistic under H0 vs t(df)",
     xlab = "t-statistic")
curve(dt(x, df = n - 1), add = TRUE)

# Empirical Type I error rate (should be close to alpha)
#mean(reject)
#table(reject)
```

# 3. Applied example: audit study (AUDIT)

## Interpretation of y
In this dataset, `y = b - w` is the difference in job-offer indicators between matched applications:
- `b = 1` if the Black applicant gets an offer, `0` otherwise
- `w = 1` if the White applicant gets an offer, `0` otherwise

So if the average difference is negative, it suggests the Black applicant is less likely to receive an offer. <-discrimination

## Test
\[
H_0:\mu_y = 0 \quad \text{vs} \quad H_1:\mu_y < 0.
\]

```{r}
summary(audit$y)
table(audit$y)

hist(audit$y, main = "AUDIT: y = b - w", xlab = "y")
```

```{r}

# Built-in
t.test(audit$y, mu = 0, alternative = "less", conf.level = 0.95)
#critical value for 1 sided 5% = -1.65
#-0.1088- 0/ se(y)
# DIY
#my_ttest(audit$y, mu0 = 0, alpha = 0.05, alternative = "less")
```

# 4. OLS in simple regression (CEOSAL1)

## 4.1 Run the regression with lm()

We estimate:
\[
salary_i = \beta_0 + \beta_1\,roe_i + u_i.
\]

```{r}
??ceosal1

plot(salary ~ roe, data = ceosal1,
     main = "CEOSAL1: Salary vs ROE",
     xlab = "ROE", ylab = "Salary")

ols_lm <- lm(salary ~ roe, data = ceosal1)
#intercept <- beta_0_hat
#roe <- beta_1_hat
summary(ols_lm)

abline(ols_lm)
```

```{r}
ols_lm2 <- lm(salary ~ sales, data = ceosal1)
summary(ols_lm2)
```

```{r}
coef(ols_lm)
coef(ols_lm2)
```

## 4.2 Replicate OLS coefficients using the formulas

The OLS slope and intercept in simple regression are:
\[
\hat\beta_1 = \frac{\sum (x_i-\bar x)(y_i-\bar y)}{\sum (x_i-\bar x)^2},
\qquad
\hat\beta_0 = \bar y - \hat\beta_1 \bar x.
\]

```{r}
my_ols <- function(x, y) {
  x <- as.numeric(x)
  y <- as.numeric(y)
  if (length(x) != length(y)) stop("x and y must have the same length.")

  xbar <- mean(x)
  ybar <- mean(y)

  dx <- x - xbar
  dy <- y - ybar  # IMPORTANT: subtract mean(y), not mean(x)

  beta1_hat <- mean(dx * dy) / mean(dx^2)
  beta0_hat <- ybar - beta1_hat * xbar

  c(beta0_hat = beta0_hat, beta1_hat = beta1_hat)
}

my_ols(ceosal1$roe, ceosal1$salary)
coef(ols_lm)
```

# 5. Simulation: sampling variation of OLS estimates (complete version)

## DGP
We simulate data from:
\[
y = \beta_0 + \beta_1 x + u,
\quad x \sim N(0,1),\ u \sim N(0,1),\ x \perp u.
\]

Then we repeat the sampling process many times and look at the distribution of $\hat\beta_0$ and $\hat\beta_1$.

```{r}
set.seed(123)

R <- 5000
n <- 100
beta0 <- 1
beta1 <- 2
beta0_hat <- numeric(R)
beta1_hat <- numeric(R)

for (r in 1:R) {
  x_sim <- rnorm(n)
  u_sim <- rnorm(n)
  y_sim <- beta0 + beta1 * x_sim + u_sim

  b <- my_ols(x_sim, y_sim)
  beta0_hat[r] <- b["beta0_hat"]
  beta1_hat[r] <- b["beta1_hat"]
}

c(mean_beta0 = mean(beta0_hat), sd_beta0 = sd(beta0_hat),
  mean_beta1 = mean(beta1_hat), sd_beta1 = sd(beta1_hat))

hist(beta1_hat, probability = TRUE,
     main = "Sampling distribution of beta1_hat",
     xlab = "beta1_hat")
curve(dnorm(x, mean = beta1, sd = sd(beta1_hat)), add = TRUE)

hist(beta0_hat, probability = TRUE,
     main = "Sampling distribution of beta0_hat",
     xlab = "beta0_hat")
curve(dnorm(x, mean = beta0, sd = sd(beta0_hat)), add = TRUE)
```

## Optional intuition: draw multiple fitted lines

```{r}
set.seed(123)

R_lines <- 30
plot(NULL, xlim = c(-3, 3), ylim = c(beta0 - 6, beta0 + 6),
     xlab = "x", ylab = "y",
     main = "Different samples -> different fitted OLS lines")

# True line
abline(a = beta0, b = beta1, lwd = 2)

for (r in 1:R_lines) {
  x_sim <- rnorm(n)
  u_sim <- rnorm(n)
  y_sim <- beta0 + beta1 * x_sim + u_sim

  b <- my_ols(x_sim, y_sim)
  abline(a = b["beta0_hat"], b = b["beta1_hat"])
}
```

# 6. Quick checklist (what you should be able to do)

- Use `t.test()` correctly for one-sided and two-sided hypotheses.
- Compute t statistics, critical values (`qt()`), and p-values (`pt()`) manually.
- Explain why the simulated rejection rate under H0 is close to \(\\alpha\).
- Run `lm()` and interpret the slope and intercept.
- Reproduce OLS coefficients using the simple-regression formulas.
- Use simulation to build intuition for sampling variability.
