---
title: "Tutorial 9: Dummy Variables and Robust Standard Errors"
date: "ECO375"
author: "Wenhui Xu"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
urlcolor: blue
geometry: left=0.75in, right=0.75in, top=0.75in, bottom=0.75in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Overview

This tutorial corresponds to the attached `tutorial_9.R`. It is concept-focused and has two goals:

1. **Dummy variables and the dummy variable trap** (perfect multicollinearity when you include all category dummies plus an intercept).
2. **Robust standard errors (White/HC)** using `sandwich::vcovHC()` and `lmtest::coeftest()`.

We use the `cps91` dataset from the `wooldridge` package.

```{r t9_packages}
if (!requireNamespace("wooldridge", quietly = TRUE)) install.packages("wooldridge")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("sandwich", quietly = TRUE)) install.packages("sandwich")
if (!requireNamespace("lmtest", quietly = TRUE)) install.packages("lmtest")

library(wooldridge)
library(dplyr)
library(sandwich)
library(lmtest)
```

# 1. Dummy variables and the dummy variable trap

## 1.1 What the problem is asking (plain English)

We want to estimate and interpret a regression of husband’s earnings (`husearns`) on union status (`husunion`, a binary variable).

Then we want to understand what happens if we include *both* dummies (`husunion` and `nounion`) and whether we include an intercept.

Key idea: if a variable is binary, the regression can be written in different (equivalent) ways—but some specifications create perfect collinearity.

## 1.2 Step-by-step plan (how to solve it)

1) Load `cps91` and inspect the binary regressor.  
2) Estimate a model with an intercept and one dummy: `husearns ~ husunion`.  
3) Create the complementary dummy `nounion = 1 - husunion`.  
4) Estimate a model **without an intercept** using both dummies: `husearns ~ 0 + nounion + husunion`.  
5) Try estimating a model **with an intercept and both dummies** and see what happens (dummy variable trap).  
6) Interpret coefficients in each specification.

## 1.3 Common pitfalls

- **Including all category dummies plus an intercept:** this creates perfect collinearity.
- **Forgetting what the intercept represents:** with a single dummy, the intercept is the mean of the base group.
- **Misreading the “no intercept” formula:** in R, `0 +` removes the intercept.

## 1.4 Data load and quick checks

```{r t9_data_load}
data("cps91")
cps91 <- as_tibble(cps91)

cps91 %>%
  summarise(
    n = n(),
    pct_union = mean(husunion == 1, na.rm = TRUE),
    mean_husearns = mean(husearns, na.rm = TRUE)
  )
```

## 1.5 Model 1: intercept + one dummy

Model:
\[
husearns = \beta_0 + \beta_1\,husunion + u
\]

Interpretation:
- \(\beta_0\): average earnings for husbands **not** in a union
- \(\beta_1\): difference in average earnings (union − nonunion)

```{r t9_lm1}
lm1 <- lm(husearns ~ husunion, data = cps91)
summary(lm1)

coef(lm1)
```

A quick way to verify the interpretation is to compute group means directly:

```{r t9_means_check}
cps91 %>%
  group_by(husunion) %>%
  summarise(
    mean_husearns = mean(husearns, na.rm = TRUE),
    n = n()
  )
```

## 1.6 Model 2: no intercept + both dummies (equivalent parameterization)

### Create the complementary dummy
```{r t9_create_nounion}
cps91 <- cps91 %>% mutate(nounion = 1 - husunion)

# sanity check: nounion + husunion should be 1 for all observations
cps91 %>% summarise(all_one = all(nounion + husunion == 1))
```

Model:
\[
husearns = \gamma_0\,nounion + \gamma_1\,husunion + u
\]
(with **no** intercept)

Interpretation:
- \(\gamma_0\): mean earnings for nonunion husbands
- \(\gamma_1\): mean earnings for union husbands

```{r t9_lm2}
lm2 <- lm(husearns ~ 0 + nounion + husunion, data = cps91)
summary(lm2)

coef(lm2)
```

## 1.7 Model 3: intercept + both dummies (dummy variable trap)

If you include an intercept **and** both dummies, you have perfect collinearity because:

\[
1 = nounion + husunion
\]

So one regressor is a perfect linear combination of the others (the intercept is a column of ones). R will drop a regressor and report `NA` for one coefficient.

```{r t9_lm3}
lm3 <- lm(husearns ~ nounion + husunion, data = cps91)
summary(lm3)
```

**Takeaway:** when you have \(G\) categories, you include:
- an intercept + \(G-1\) dummies, **or**
- no intercept + all \(G\) dummies.

# 2. Robust standard errors (White / HC)

## 2.1 What the problem is asking (plain English)

The usual OLS standard errors assume homoskedasticity (constant variance of \(u\) given \(x\)).

If heteroskedasticity is present, OLS coefficients are still unbiased/consistent (under the usual assumptions), but the **usual** standard errors can be wrong. Robust (White/HC) standard errors fix the standard error calculation.

In this tutorial, we compute robust standard errors for `lm1`.

## 2.2 Step-by-step plan (how to solve it)

1) Fit the OLS model `lm1`.  
2) Compute a robust variance-covariance matrix using `sandwich::vcovHC(lm1, type="HC0")`.  
3) Extract robust standard errors via `sqrt(diag(vcov_robust))`.  
4) Use `lmtest::coeftest()` to print coefficient tests using robust SE.  
5) (HW) Compute the robust variance matrix “by hand” using the sandwich formula and verify it matches `vcovHC()`.

## 2.3 Common pitfalls

- **Forgetting that robust SE change t statistics and p-values but not coefficients.**
- **Using the wrong HC type:** HC0 is the original White; HC1 is a small-sample correction (often a good default in practice).
- **Mixing up variance and standard error:** SE is the square root of the variance (diagonal element).

## 2.4 Module: What are `vcovHC()` and `coeftest()`? (first-time tools)

### `sandwich::vcovHC()`
`vcovHC()` computes a **heteroskedasticity-consistent** covariance matrix for the OLS coefficients.

The generic White “sandwich” estimator is:
\[
\widehat{\mathrm{Var}}(\hat\beta)
= (X'X)^{-1} \left( \sum_{i=1}^n \hat u_i^2 x_i x_i' \right) (X'X)^{-1}.
\]

- `type="HC0"` corresponds closely to the expression above.
- Other types (HC1, HC2, HC3, …) make small-sample or leverage adjustments.

### `lmtest::coeftest()`
`coeftest()` prints coefficient tests and allows you to **plug in a custom covariance matrix** (like a robust one).

That means you can keep the same OLS coefficients but replace the standard errors used for inference.

## 2.5 Robust SE using `vcovHC()` + `coeftest()`

```{r t9_robust_vcov}
robust_vcov <- vcovHC(lm1, type = "HC0")

# Robust standard errors for each coefficient
sqrt(diag(robust_vcov))
```

```{r t9_coeftest_robust}
# Use robust VCOV inside coeftest.
# df = Inf uses the large-sample (normal) reference distribution.
coeftest(lm1, df = Inf, vcov = robust_vcov)
```

## 2.6 HW (complete): compute White (HC0) robust SE “by hand”

We will manually construct the sandwich estimator:

\[
\widehat{\mathrm{Var}}(\hat\beta)
= (X'X)^{-1} \, X' \widehat{\Omega} X \, (X'X)^{-1},
\quad \widehat{\Omega} = \mathrm{diag}(\hat u_1^2,\ldots,\hat u_n^2).
\]

```{r t9_manual_hc0}
# Design matrix X and residuals uhat
X <- model.matrix(lm1)     # includes intercept column automatically
uhat <- resid(lm1)

# Sandwich pieces
XtX_inv <- solve(t(X) %*% X)
meat <- t(X) %*% (diag(uhat^2)) %*% X

vcov_hc0_manual <- XtX_inv %*% meat %*% XtX_inv

# Compare manual vs vcovHC
max_abs_diff <- max(abs(vcov_hc0_manual - robust_vcov))
max_abs_diff
```

Robust SE comparison (manual vs `vcovHC`):

```{r t9_manual_vs_vcovhc}
cbind(
  se_manual = sqrt(diag(vcov_hc0_manual)),
  se_vcovHC = sqrt(diag(robust_vcov))
)
```

# Wrap-up

- Dummy variables: choose a base category (intercept + one dummy) OR remove intercept and include all category dummies.
- Dummy variable trap: intercept + all dummies creates perfect collinearity.
- Robust SE: `vcovHC()` changes standard errors (and inference), not the OLS coefficient estimates.
- `coeftest()` is a clean way to report robust t tests without rewriting the regression output.
